---
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{tcolorbox}
- \usepackage{fancyhdr}
- \usepackage[utf8]{inputenc}
- \usepackage{wrapfig}
- \usepackage{amsmath}
- \usepackage{booktabs}
- \usepackage{esvect}
urlcolor: blue
---

\pagestyle{fancy}
\fancyhf{}
\rhead{Will Nickols}
\lhead{Section 1}
\rfoot{Page \thepage}

# Announcements

\begin{wrapfigure}{r}{0.12\textwidth}
  \centering
    \vspace*{-1.3cm}
    \includegraphics[width=\linewidth]{section_qr_code.png}
\end{wrapfigure}

Make sure to sign in on the [google form](https://forms.gle/xm1DfzuZFNcWU6fH8) (I send a list of which section questions are useful for which pset questions afterwards)

Pset 0 due Friday 9/15

# Introductions
- Name
- Year
- Previous stats courses
- One question or thought related to lecture last week

# Goals each week
- Hand out and explain R code for the week. New relative to last year, we'll plan to not do any in-section coding questions. LLMs are good enough now to do most of your coding for you (and they're allowed in this class!).
- See similar examples to the homework (both in code and analysis).
- Learn something about the world.

```{r, echo=F, warning=F, message=F, cache=F}
list.of.packages <- c("ggplot2", "gridExtra", "MASS")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(ggplot2)
library(gridExtra)
library(MASS)
```

# Effective sample size

The following problems are intended as a review of Stat 110.

1. Suppose there is a gambler who goes to the casino for $n$ days and makes $Z_1, Z_2, \dots, Z_n\sim\mathcal{N}(0, 1)$ each day where the winnings are independent of each other. (You can assume these are in thousands if the stakes aren't high enough.) What is the distribution of $\bar{Z}$?

\vspace{0.5 cm}

2. Now, suppose the gambler tends to win and lose in streaks. In particular, let $X_1, X_2, \dots, X_n\sim\mathcal{N}(0, 1)$ marginally be the winnings, but assume neighboring days have correlation $\rho$. That is, $$\vv{X}\sim\textrm{MVN}(\vv{0},\mathbf{\Sigma}), \mathbf{\Sigma}=
\begin{bmatrix} 
    1 & \rho & 0 & 0 & \dots \\
    \rho & 1 & \rho & 0 & \dots \\
    0 & \rho & 1 & \rho & \dots \\
    0 & 0 & \rho & 1 & \dots \\
    \vdots & \vdots &\vdots &\vdots &\ddots \\
\end{bmatrix}$$ Intuitively, should the variance of $\bar{X}$ be higher or lower than the variance of $\bar{Z}$?
    
\vspace{1 cm}

3. What is the distribution of $\bar{X}$?

\vspace{6 cm}

4. What would the distribution be if the $X_i$ had variance $\sigma^2$ instead of $1$ but everything else remained the same?

\vspace{2 cm}

5. What is the approximate distribution for large $n$?

\vspace{2 cm}

6. By comparing the distributions in (1) and (5), determine the effective sample size $n'$ when there are $n$ random variables with the correlation structure of (2). That is, is you had $n'$ independent Normals rather than $n$ dependent Normals, what would $n'$ have to be so that the variances of the sample means are the same?

\vspace{2 cm}

7. Here is a plot of how the effective sample size changes with $\rho$.

```{r, echo=F, fig.width=4, fig.height=3, fig.align='center', cache=T}
rhos <- seq(0, 0.5, 0.001)
eff_samp_size <- function(rhos) {
  return(1/(1 + 2 * rhos))
}
ggplot(data.frame(x=rhos), aes(x=x)) + 
  geom_function(fun = eff_samp_size, colour = "black", linewidth=2) + 
  ylim(c(0, 1)) + 
  theme_bw() + 
  xlab("Correlation") + 
  ylab("n'/n")
```

We can test that our calculations are right by using a simulation. Explain what the following code does and whether the results agree with our expectations.

```{r, cache=T}
library(MASS) # For Multivariate Normal
set.seed(139)

nsim <- 10^5
n <- 70
p <- 0.2
n_eff <- as.integer(n / (1 + 2 * p))

Sigma = matrix(0, nrow = n, ncol = n)
diag(Sigma) <- 1
for (i in 2:n) {
  Sigma[i, i-1] <- p
  Sigma[i-1, i] <- p
}

outputs <- matrix(nrow = nsim, ncol = 3)
for (i in 1:nsim) {
  x <- rnorm(n, 0, 1)
  outputs[i,1] <- mean(x)
  
  x <- rnorm(n_eff, 0, 1)
  outputs[i,2] <- mean(x)
  
  x <- mvrnorm(n = 1, rep(0, n), Sigma)
  outputs[i,3] <- mean(x)
}

variances_out <- apply(outputs, 2, var) # Apply over columns
names(variances_out) <- c("Independent n", "Independent n'", "Dependent n")
variances_out
```

\vspace{3 cm}

8. You might have noticed that the plot of effective sample size versus correlation stops at a correlation of 0.5. Correlation ranges from -1 to 1, but our set-up actually doesn't work if $\rho>0.5$ and $n$ is large enough. To have a valid $\mathbf{\Sigma}$ matrix, it must satisfy the property that $\vv{x}^T\mathbf{\Sigma} \vv{x}\geq 0$ for all $\vv{x}\in\mathbb{R}^n$ (that is, it must be positive, semi-definite). Show that for $\rho>0.5$, choosing the vector $\vv{x}=(-1, 1, -1, ..., -1)^T$ implies $\vv{x}^T\mathbf{\Sigma} \vv{x}<0$ if $n$ is large enough, violating the requirements for $\mathbf{\Sigma}$. (For simplicity, let $n$ be odd.)

\vspace{6 cm}

\newpage

# Student-$t$ vs Normal

The following problems are intended as a review of Stat 111. We'll prove that the student-$t$ distribution converges to the Normal distribution as its degrees of freedom increase and then analyze this convergence. This fact is useful for large $n$ approximations.

1. Let $T_n\sim t_n$, so $T_n$ can be represented as $$T_n=\frac{Z}{\sqrt{V/n}}, Z\sim\mathcal{N}(0,1), V\sim\chi^2_n$$ which also means $V$ can be represented as $V=\sum_{i=1}^n Z_i^2$ for $Z_i\sim\mathcal{N}(0,1)$. Show that $V/n\xrightarrow{p}1$.

\vspace{2 cm}

2. What tells us that if $V/n\xrightarrow{p} 1$, $\frac{1}{\sqrt{V/n}}\xrightarrow{p} 1$?

\vspace{0.5 cm}

3. What tells us that if $Z\sim\mathcal{N}(0,1)$ and $\frac{1}{\sqrt{V/n}}\xrightarrow{p} 1$, $\frac{Z}{\sqrt{V/n}}\xrightarrow{d} \mathcal{N}(0,1)$

\vspace{0.5 cm}

4. What does this mean about the distribution of $T_n$ as $n\rightarrow\infty$?

\vspace{0.5 cm}

5. Do the centers or the tails converge faster?

```{r, echo=F, fig.width=6, fig.height=3, fig.align='center', cache=T}
dfs = seq(1, 30, 0.001)
ratios <- function(dfs, point) {
  return((pt(point, dfs, lower.tail = F)) / pnorm(point, lower.tail = F))
}
ggplot(data.frame(x=dfs), aes(x=x)) + 
  geom_function(fun = ratios, aes(color = "0"), args = list(point = 0)) +
  geom_function(fun = ratios, aes(color = "1"), args = list(point = 1)) +
  geom_function(fun = ratios, aes(color = "2"), args = list(point = 2)) +
  geom_function(fun = ratios, aes(color = "2.5"), args = list(point = 2.5)) +
  theme_bw() + 
  scale_color_manual(values = c("0" = "black", "1" = "darkred", "2" = "red", "2.5" = "pink")) +
  labs(color = "Tail mass beyond: ") + 
  xlab("Degrees of freedom") + 
  xlim(c(1,30)) + 
  ylim(c(0,20)) + 
  ylab("t tail mass/Normal tail mass")
```

\vspace{1 cm}

6. What does this imply about generating p-values from a Normal approximation to the student-$t$ distribution?

\vspace{3 cm}

\newpage

# Country demographics

These problems will deal with a data set of country-level statistics from [UNdata](https://data.un.org/) and [Varieties of Democracy](https://v-dem.net/data/the-v-dem-dataset/).

```{r, echo=F}
# Read in the data
countries <- read.csv("data/country_stats.csv", check.names = F)
```

1. Compare the following summary statistics for the 2010 populations (in millions of people) of Western African and Eastern African countries:

```{r}
# Western Africa
pop1 <- countries[countries$Year == 2010 & 
                    countries$Region == "Western Africa",
                  ]$`Population mid-year estimates (millions)`
round(c(summary(pop1), "SD" = sd(pop1)), 2)

# Eastern Africa
pop2 <- countries[countries$Year == 2010 & 
                    countries$Region == "Eastern Africa",
                  ]$`Population mid-year estimates (millions)`
round(c(summary(pop2), "SD" = sd(pop2)), 2)
```

\vspace{3 cm}

2. Compare the distributions. Would you expect to see a significant difference in a $t$-test?

```{r, echo=F, fig.width=6, fig.height=3, fig.align='center', cache=T}
country_subset <- countries[countries$Year == 2010 & 
                              countries$Region %in% c("Western Africa", "Eastern Africa"),]
plot1 <- ggplot(country_subset, aes(x=`Population mid-year estimates (millions)`, fill=Region)) + 
  geom_histogram(alpha=0.4, position="identity", bins = 30) + 
  xlab("Population (millions)") + 
  ylab("Countries") + 
  theme_bw() + 
  theme(legend.position = "none")

plot2 <- ggplot(country_subset, aes(x=Region,y=`Population mid-year estimates (millions)`, fill=Region)) + 
  geom_boxplot(alpha=0.4, outlier.size = 0) + 
  xlab("Population (millions)") + 
  ylab("Region") + 
  theme_bw() + 
  geom_jitter(color="black", size=1, alpha=0.9, width = 0.1) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

grid.arrange(plot1, plot2, ncol=2, widths=c(1,1.2))
```

\vspace{2 cm}

3. Perform a formal $t$-test for the difference in population means between Western African and Eastern African countries.

```{r, echo=F}
west_african <- pop1
east_african <- pop2
t.test(west_african, east_african)
```

\vspace{1 cm}

4. Perform a formal $z$-test for the difference in the proportions of the populations that are nurses or midwives in the US versus the UK in 2010.

```{r, echo=F}
us_nurses_midwives <- countries$`Health personnel: Nurses and midwives (number)`[
  countries$Country == "United States of America" & countries$Year ==  2010]
us_pop <- countries$`Population mid-year estimates (millions)`[
  countries$Country == "United States of America" & countries$Year ==  2010] * 1000000
uk_nurses_midwives <- countries$`Health personnel: Nurses and midwives (number)`[
  countries$Country == "United Kingdom" & countries$Year ==  2010]
uk_pop <- countries$`Population mid-year estimates (millions)`[
  countries$Country == "United Kingdom" & countries$Year ==  2010] * 1000000
prop.test(c(us_nurses_midwives, uk_nurses_midwives), c(us_pop, uk_pop), correct = F)
```

\vspace{1 cm}

5. Varieties of Democracy is a group of researchers that estimates a democracy score for each country each year based on a large compilation of data. Note any trends in the democracy index.

```{r, echo=F, fig.width=6, fig.height=3, fig.align='center', cache=T, warning=F}
ggplot(countries[countries$Country %in% c("United States of America",
                                          "Russian Federation"),], aes(x=Year, y=v2x_polyarchy, color=Country)) + 
  geom_point(size=2) + 
  ylim(c(0,1)) + 
  xlab("Year") + 
  ylab("Democracy index (0-1)") + 
  theme_bw()
```

\vspace{3 cm}
