---
output: pdf_document
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{fancyhdr}
   - \usepackage[utf8]{inputenc}
---

\pagestyle{fancy}
\fancyhf{}
\rhead{Will Nickols}
\lhead{Section 6}
\rfoot{Page \thepage}

# Announcements
- Make sure to sign in on the [google form (linked here)](https://forms.gle/JGvZP8CPUhaefnLT6)
- Pset 5 due October 21 at 5 pm

# Normal interactions

Let $Z\sim\mathcal{N}(0,1)$ and $X\sim\mathcal{N}(\mu,\sigma^2)$.

1. Show that $\textrm{Corr}(Z, Z^n)=0$ for all even whole numbers $n$.


2. Show that $\textrm{Corr}(Z, Z^n)>0$ for all odd whole numbers $n$.  You may use the useful fact from the Stat 110 book (page 284) that $E(Z^{2n})=\frac{(2n)!}{2^nn!}$ for integers $n\geq 0$.

3. Find $\textrm{Cov}(X, X^2)$.  When will this be positive?  When will this be negative?  (Hint: Consider standardizing $X$.)

4. What implication does this have for fitting linear models with a Normal predictor and its squared term?

# Island of Misfit Toys

This section will deal with a data set of country-level statistics from [this source](https://www.gu.se/en/quality-government/qog-data/data-downloads/standard-dataset) with an explanation of the data encoding found [here](https://www.qogdata.pol.gu.se/data/codebook_std_jan22.pdf).

```{r, cache=T, warning=FALSE}
countries <- read.csv("data/countries.csv")
```

A few useful columns:

- `mad_gdppc`: GDP per capita
- `ht_region`: Country's region of the world: Eastern Europe (1), Latin America (2), North Africa & the Middle East (3), Sub-Saharan Africa (4), Western Europe and North America (5), East Asia (6), South-East Asia (7), South Asia (8), Pacific (9), Caribbean (10)
- `wdi_araland`: Arable land (\% of land area)
- `wdi_precip`: Average annual precipitation (mm per year)
- `spi_ospi`: Overall social progress index on 0-100 scale
- `bmr_dem`: Binary democracy measure

1. Using `relevel` to set Western Europe and North America as the reference group, fit a regression model to predict a country's GDP per capita from its region.  Interpret the coefficients.

2. Build a 2nd order polynomial regression model to predict the proportion of arable land in a country from its average annual precipitation.  Interpret the output and provide a visual.

```{r, cache=T, warning=FALSE}
library(ggplot2)

# TODO: Get summary of 2nd order polynomial model

ggplot(countries, aes(x=wdi_precip, y=wdi_araland)) + 
  geom_point() + 
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 2)) + 
  ylim(0, 60) + 
  theme_bw()
```

3. Use the previous model to find the probability that a country with 1500 mm annual precipitation per year on average will have less than 10\% of its land arable.  

```{r, include=F, cache=TRUE, eval=F}
# TODO: Find probability

# Consider this starter code that finds the density under -7.61 and under 44.16
predict(lm_fit,newdata = data.frame(wdi_precip=1500),interval = "prediction")
threshold = c(-7.61454, 44.1593)

pt((threshold - as.vector(C %*% coef(lm_fit))) / 
     as.vector(sqrt(t(C) %*% vcov(lm_fit) %*% C + summary(lm_fit)$sigma^2)), 
   lm_fit$df.residual)
```

4. Fit a LOESS model for the same data and compare its prediction accuracy to that of the previous model.

```{r, cache=T, warning=FALSE}
# TODO: Fit LOESS model

# TODO: LM Rsq

# TODO: Loess Rsq

ggplot(countries, aes(x=wdi_precip, y=wdi_araland)) + 
  geom_point() + 
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 2)) + 
  stat_smooth(method="loess", 
              formula = y ~ x,
              col="red") + 
  ylim(0, 60) + 
  theme_bw()
```

5. Fit a model to predict a country's overall social progress index from the log of its GDP per capita, its democracy status, and the interaction of the two.  Interpret the coefficients of the model.

```{r, cache=T, warning=FALSE}
# TODO: Fit model

ggplot(countries, aes(x=log(mad_gdppc, 2), y = spi_ospi, col = as.factor(bmr_dem))) + 
  geom_point() + 
  stat_smooth(method="lm",
              formula = y~x)
```

6. Perform a formal hypothesis test to determine whether the previous model performs significantly better at predicting the overall social progress index than a model without the interaction term.

```{r, cache=T, warning=FALSE}
# TODO: Fit model and perform a test
```

# Everything is just a linear model

Let $Y_{ij}$ be data point $j$ from group $i$ where there are $k$ groups with $n_i$ data points in group $i$.  Imagine we run an ANOVA as well as an $F$-test for overall significance of a regression model with only the categories as predictors.  Recall the original ANOVA $F$-statistic: $$\frac{\sum_{i=1}^kn_i(\bar{Y}_i-\bar{Y})^2/(k-1)}{\sum_{i=1}^k\sum_{j=1}^{n_i}(Y_{ij}-\bar{Y_i})^2/(n-k)}$$ and the overall regression $F$-statistic: $$\frac{\sum_{i,j}(\hat{Y}_{ij}-\bar{Y})^2/p}{\sum_{i,j}(Y_{ij}-\hat{Y}_{ij})^2/(n-p-1)}$$ where $p$ is the number of predictors (not including the intercept in the model).

1. What is $p$ in this case?

2. What is $\hat{Y}_{ij}$?  Why is this the case?

3. Show that the two $F$-statistics are equal.

# Simpson's simulation

1. For the following data table, write out the design matrix that would be used in the following model: `response ~ category * value`.

```{r, cache=T, warning=FALSE, echo=F}
set.seed(139)
library(knitr)
library(dplyr)

n = 10
value = rnorm(n, 5, 2)
category = as.factor(sample(c(1,2,3), n, replace = T))
df <- data.frame(Response = case_when(category == "1" ~ 3,
                                      category == "2" ~ 7,
                                      category == "3" ~ 2) * value + rnorm(n, 0, 5),
                 Category=category,
                 Value=value)

kable(df)
```

2. Without looking at the code that generated the data, for each of the pairs of plots below, determine what model should be fit to best describe the data.

```{r, echo=F, cache=T, warning=F, fig.height=3.5}
library(gridExtra)

# Sample size
n <- 1000

# Five categories
category <- sample(1:5, n, replace = T)

# Different x mean for each category
x <- rnorm(n, 10 * category, 15)

# Generate the response variable
response = 100 - (40 * category) + x * 1 + rnorm(n, 0, 15)

# Make a dataframe for plotting
df <- data.frame(x = x,
                 category = as.factor(category),
                 response = response)

# Models with and without category as a predictor
plot1 <- ggplot(df, aes(x = x, y = response)) + 
  geom_point() + 
  theme_bw() + 
  geom_smooth(method="lm", formula = y~x)

plot2 <- ggplot(df, aes(x = x, y = response, color = category)) + 
  geom_point() + 
  theme_bw() + 
  geom_smooth(method="lm", formula = y~x)

# Plot models
grid.arrange(plot1, plot2, ncol = 2, widths=c(0.8,1))
```

```{r, echo=F, cache=T, warning=F, fig.height=3.5}
n <- 1000

# Two categories
category <- sample(0:1, n, replace = T)

# Generate predictor
x <- rnorm(n, 0, 15)

# Generate response variable
response = x * (-1 + 2 * category) + rnorm(n, 0, 15)

# Data frame for plotting
df <- data.frame(x = x,
                 category = as.factor(category),
                 response = response)

# Models with and without interaction terms
plot1 <- ggplot(df, aes(x = x, y = response)) + 
  geom_point() + 
  theme_bw() + 
  geom_smooth(method="lm", formula = y~x)

plot2 <- ggplot(df, aes(x = x, y = response, color = category)) + 
  geom_point() + 
  theme_bw() + 
  geom_smooth(method="lm", formula = y~x)

grid.arrange(plot1, plot2, ncol = 2, widths=c(0.8,1))
```

3. Name a reason to avoid fitting many interaction terms right from the beginning.

